---
title: "Using interactive tutorials"
author: "Nathan L. Brouwer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The abdcompanion package contains interactive tutorials for key examples from Whitlock and Schluter's texbook *Analysis of Biological Data*.  There are two general groups of tutorials

1. Interactive web-based tutorials 
1. Text-based tutorials

The latter, text-based tutorials are what you will find in most books teaching R or other software.  They provide step by step written instructions that you can work through by typing commands into the R console or executing them from an R script.  To use these tutorials you will have to have a basic familiarity with R, and have to figure out how to trouble shoot anything that comes up.  The rest of the "vignettes" of the abdcompanion package are these conventinal R tutorials.

The interactive tutorials are much more dynamic, and require almost no knowledge of R.  After loading the desired tutorial in R, your web browser will open up an interative screen.  All data will be shown in tables and you will be lead through any computations or R code that needs to be written.

These interactive tutorials are further divided into 2 gorups

* Spreadsheet-oriented tutorials
* R tutorials

### Interactive "spreadsheet"" tutorials 

These "spreadsheet"" tutorials are meant to be either  worked through on their own or, ideally, with a spreadsheet such as Excel or Google Sheets open.  R does not have much real spreadsheet capabilities, and but the general output you should be creating is displayed.  An imporant part of these tutorials are prompts to answer questions about the calculations that are being demonstrated (and which you are hopefully doing in a spreadsheet as you go!).   At times there will be small windows where you can input numbers like a scientific calculator to do simple calculations.  Links are provided to example spreadsheets and worksheets that can be filled out based on the calculations being done.

<br>

### Interative R Tutorials

The interactive R-tutorials are designed to help you learn the basics of R while providing a lot of help.  Like the "spreadsheet" exercises, these tutorials don't run in R (or a spreadhseet) but in your web browser.  Aspects of an R session are emulated and you can type code into input windows and see the output.  Importantly, you can get hints and help to figure out the code you need to write.  There are also questiosn to test you understanding as you go.


## Running the interactive tutorials

To use the interactive tutorials you will need

* R to be installed on your computer
* RStudio to be installed
* The devtools and learnr packages from CRAN
* The abdcompanion package


### Getting the devtools package 

To get devtools installed, you can type 

```{r eval = F}
install.packages("devtools")
```

Or find the Files / Plots / Packages / Help / Viewer tab in RStudio, click on Packages, and then Install, and enter "devtools."

Do the same thing to get the learnr package.

### Getting the abdcompanin package from GitHub

To install the abdcompanion package, first download it from GitHub using the install_github() function from the devtools package.
```{r eval = F}
devtools::install_github("brouwern/abdcompanion")
```

Now explicitly load the packge using the library() function.
```{r eval = F}
library(abdcompanion)
```


### Running an interactive tutorial

Finally, run the desired tutorial using the run_tutorial() function from the learnr package.

NOTE: it is called "toad", no "s";
```{r eval = FALSE}
learnr::run_tutorial("toad_spreadsheet", package = "abdcompanion")
```

R will run some code in the console (eg "Listening on http://127.0.0.1:4051"), and your default web browser will open up.  You can then click through the tutorial.

### Closing 

When you are done with the tutorial close down your web browser.  R will still be in tutorial mode.  Click on the red stopsign icon in the upper right hand side of the the console to close this down.  You can then use R normally.
